### 2022-08-23 main/chapter_2ljets.tex
more probability posteriors etc, ref cartoon

Worse, I disregarded the popular idea that every region should have a large
background expectation, preferably of ten events or more.
This requirement is typically justified by the limitations of approximate
statistical methods that work
``from as few as $\mathcal{O}(10)$ data events''~\cite{histfitter2014}.
But the Poisson likelihood works for infinitesimally small
expectations~\cite{skilling2009cosmology}, and in the small-bin limit one has
continuous of `unbinned' distributions with Poisson normalizations
(sometimes called ``extended'' in high-energy
physics~\cite{barlow1990extended})
that are standard elsewhere in high-energy physics,
and are easily analysed by the standard likelihood fitting
methods~\cite{cowan1998statistical}.
Examples uses of unbinned modelling include analyses of hadronic
resonances in
\lhcb~\cite{Tullythesis,lhcb2018563}, searches for exotic effects in
\atlas~\cite{EXOT-2018-08},
and notably the observation of the
Higgs boson~\cite{HIGG-2012-27}.

>>>>

Worse, I disregarded the popular idea that every region should have a large
background expectation, preferably of ten events or more.
This requirement is typically justified by the limitations of approximate
statistical methods that work
``from as few as $\mathcal{O}(10)$ data events''~\cite{histfitter2014}.
But the Poisson likelihood works for infinitesimally small
expectations~\cite{skilling2009cosmology}, and in the small-bin limit one has
continuous of `unbinned' distributions with Poisson normalizations
(sometimes called ``extended'' in High-Energy
Physics~\cite{barlow1990extended})
that are standard elsewhere in High-Energy Physics,
and are easily analysed by the standard likelihood fitting
methods~\cite{cowan1998statistical}.
Examples uses of unbinned modelling include analyses of hadronic
resonances in
\lhcb~\cite{Tullythesis,lhcb2018563}, searches for exotic effects in
\atlas~\cite{EXOT-2018-08},
and notably the observation of the
Higgs boson~\cite{HIGG-2012-27}.

### 2022-08-23 main/chapter_searches.tex
more probability posteriors etc, ref cartoon

When we fill a histogram by summing weights of simulated samples, we are
calculating another likelihood over hypotheses $h$ alone, not $`x, h\textrm{'}$
jointly.
To disambiguate the two, we call this abstracted, $x$-independent likelihood
the \textbf{evidence} $Z(h)$ for hypothesis $h$.
The evidence is calculable as
\begin{align}
Z
= \prob{d}{h} &= \sum_i L(x_i) \,\pi(x_i) \\
&= \int\! L(x) \,\mathrm{d}\pi(x),
\label{eqn:searches_evidence_integral_measure}
\end{align}
where Equation~\ref{eqn:searches_evidence_integral_measure} is notation for
integrating the likelihood function over the prior measure~\cite{
billingsley2008probability,
skilling2010foundations
}, and relates most closely to our practice of simulating from the model and
counting events that hit the non-zero likelihood in the bin.
We shall shortly derive this identity.
The name `evidence' relates to its application for model comparison.
Supposing we have this one event with $\met > 100\,\eV[G]$, did it come from
the background or signal sample?
The answer depends on the signal and background cross-sections in our collider,
but larger evidence for signal does influence any conclusion to move in its
favour.

@@

Posterior probabilities are, however, useful when digesting data to make
inferences about the world.
Reporting likelihoods and evidence values allows readers to assign priors as
they wish.
For model comparison between background $h=B$ and signal $h=S$, this is clearly
expressed in the ratio form
% likelihood ratio
\begin{equation}
\hphantom{\quad \quad \mid\mid \mathrm{context}}
\frac{\prob{S}{d}}{\prob{B}{d}}
=
\!\!
\frac{\prob{d}{S}}{\prob{d}{B}}
\!\!\times~
\frac{\probc{S}}{\probc{B}}
\quad \quad \mid\mid \mathrm{context}
,
\end{equation}
where we use the double bar, $\mid\mid$, to refer to any contextual information
a reader might wish to include.
Whenever the evidence ratio is independent of that context, it factors out from
the prior (odds) ratio.
In this sense, likelihood (or evidence) ratios are scale factors by which prior
odds are updated.

The infamy of Bayes' rule stems from a fixation on the posterior distributions
from unjustified prior assignments.
Classical works by Bayes and Laplace~\cite{bayes1763lii, laplace1774stigler}
began this focus;
both assume prior densities that are uniform in their chosen coordinates, a
practice that can yield practical results in posterior inferences.
But for our purposes of reporting scientific data, such practices are fairly
criticized as arbitrary when they influce our results.
\TODO{references for definition of Bayesian analysis as computation of
posterior without even giving a symbool to the evidence}
Figure~\ref{fig:searches_baton_roue_bayes}

\TODO{zero evidence because for uniform on the real line --- a normalized prior
with any inkling of the scale of the actual parameter will be infinitely
preferred in its evidence.}
oceeds as follows:
\begin{enumerate}
\item Predict new data $d$ with probability $\prob{d}{\ldots, h}$.
\item Update your predictions with the observed data to $\prob{x}{d, \ldots, h}$.
\item Go to 1.
\end{enumerate}

% bayes
\TODO{move to connection to frequentist section}
Bayes' theorem is an infamous and necessary result, and it is not seriously
disputed. What has been disputed is whether probability can be applied to
logic, or any application other than the random happenings of events.

% posterior

\paragraph{Probability assignments.}
infamously arbitrary and subjective. This is reality. They are.
Consider this simple mathematical example with two propositions
$a=\textrm{``}x\textrm{ is even''}$ and
$b=\textrm{``}x\textrm{ equals }2\textrm{''}$.
Clearly $\prob{a}{b} = 1$ since $2$ is indeed even.
But what is $\prob{b}{a}$? We don't know. Perhaps one could argue symmetry
across all even numbers, but there are infinitely many of them so that
probability distribution could not be normalized.

\TODO{\ldots}

Succinctly stated by Jaynes that
``any function with a single rounded maximum, raised to a higher and higher
power, goes into a Gaussian function''~\cite{jaynes2003probability}.

@@

>>>>

When we fill a histogram by summing weights of simulated samples, we are
calculating another likelihood over hypotheses $h$ alone, not $`x, h\textrm{'}$
jointly.
To disambiguate the two, we call this abstracted, $x$-independent likelihood
the \textbf{evidence} $Z(h)$ for hypothesis $h$.
The evidence is calculable as
\begin{align}
Z
= \prob{d}{h} &= \sum_i L(x_i) \,\pi(x_i)
\label{eqn:searches_evidence_integral_sum}
\\
&= \int\! L(x) \,\mathrm{d}\pi(x),
\label{eqn:searches_evidence_integral_measure}
\end{align}
where Equation~\ref{eqn:searches_evidence_integral_measure} is notation for
integrating the likelihood function over the prior measure~\cite{
billingsley2008probability,
skilling2010foundations
}, and relates most closely to our practice of simulating from the model and
counting events that hit the non-zero likelihood in the bin.
We shall shortly derive this identity.
The name `evidence' relates to its application for model
comparison~\cite{mackay2003information}.
Supposing we have this one event with $\met > 100\,\eV[G]$, did it come from
the background or signal sample?
The answer depends on the signal and background cross-sections in our collider,
but larger evidence for signal does influence any conclusion to move in its
favour.

@@

Posterior probabilities are, however, useful when digesting data to make
inferences about the world.
Reporting likelihoods and evidence values allows readers to assign priors as
they wish.
For model comparison between background $h=B$ and signal $h=S$, this is clearly
expressed in the ratio form
% likelihood ratio
\begin{equation}
\hphantom{\quad \quad \mid\mid \mathrm{context}}
\frac{\prob{S}{d}}{\prob{B}{d}}
=
\!\!
\frac{\prob{d}{S}}{\prob{d}{B}}
\!\!\times~
\frac{\probc{S}}{\probc{B}}
\quad \quad \mid\mid \mathrm{context}
,
\end{equation}
where we use the double bar, $\mid\mid$, to refer to any contextual information
a reader might wish to include.
Whenever the evidence ratio is independent of that context, it factors out from
the prior (odds) ratio.
In this sense, likelihood (or evidence) ratios are scale factors that update
any odds ratio by the same multiplication.

The infamy of Bayes' rule stems from a fixation on the posterior distributions
from unjustified prior assignments.
Classical works by
Bayes~\cite{bayes1763lii} and
Laplace~\cite{laplace1774stigler} began this focus ---
both assume prior densities that are uniform in chosen coordinates, a practice
that can yield practical results in posterior inferences.
But to report scientific data, such practices are fairly criticized as
arbitrary when they influce the reported results.
Prior assignments are arbitrary.
You are welcome to propose any prediction in any circumstances, and the
scientific response is to test whether or not its predictions are compatible
with our observations of nature.
Our emphasis on the likelihood functions and evidence values that quantify
those predictions avoids the pitfall of seeking uninformative, neutral, or
unbiased priors by embracing priors as active and explicit predictions.
This is the modern, practical approach to Bayesian data analysis~\cite{
mackay2003information,
skilling2004nested,
skilling2006nested,
sivia2006data,
skilling2010foundations
}.

Some sources continue to define Bayesian analysis as the computation of
posterior probabilities given data, often without naming the evidence, and
expressing it only as the sum from
Equation~\ref{eqn:searches_evidence_integral_sum}
--- as just a normalizing constant~\cite{
gelman1995bayesian,
gelman2008objections,
DAgostini:1994fjx,
DAgostini:2010hil,
cowan1998statistical,
pdg2020review
}.
Bayes' rule has two outputs, and both outputs have uses.
Cutting out the evidence hamstrings Bayesian analysis, leaving only the
semi-Bayesian posterior-only approach and its clear flaws.
A bicycle is stable with two wheels and falls over if one either
is removed;
Figure~\ref{fig:searches_baton_roue_bayes} illustrates this analogy.
Although it is possible to ride a unicycle, it is a circus trick and not often
a practical tool.

The essence of this form of Probabilistic data analysis is most succinctly
summarized by Skilling~\cite{skilling2008rant}:
``we SHOULD use Bayes by assigning subjective priors and inspecting evidence
values.''
This agrees exactly with the standard practice of histogramming simulations,
as illustrated in Figure~\ref{fig:searches_sig_bkg_prior_likelihood}.
Signal and background models define priors, and their (normalized) bin yields
are the evidence values that we use to compare the models.
Finally corresponding posteriors are the normalized shapes those samples take
within the observed histogram bin, which we do not use to compare the models,
but which could in principle be used for secondary interpretations.

\paragraph{Probability assignments.}
Probability algebra manipulates probabilities, but does not state how they
should be initially assigned.
Probability assignments can, in general, be freely assigned and refined from
calibration data. But there are unambiguous cases that are commonly used in
High-Energy, notably the Poisson and Gaussian distributions.
To avoid turning this thesis into a textbook, we only sketch the derivations
of these, and refer to real textbooks for their detailed
derivations~\cite{jaynes2003probability}.

Both Poisson and
Binary accept/reject experiments are described by the
The Large Hadron Collider creates $40$~million proton collisions per second,
only around $1000$ are accepted by \atlas' trigger system per
second~\cite{atlas2020trigger}.
This acceptance rate of $250$ parts per million is sufficient to invoke the
Poisson distribution for event counts.
Each collision event has some probability $p$ of passing the trigger
\begin{equation}
\label{eqn:searches_poisson}
\mathrm{Poisson}(n\mid \lambda) = \frac{\lambda^n}{n!}\exp{(-\lambda)}
\end{equation}

Succinctly stated by Jaynes that
``any function with a single rounded maximum, raised to a higher and higher
power, goes into a Gaussian function''~\cite{jaynes2003probability}.
\begin{equation}
\label{eqn:searches_gaussian}
\mathrm{Gaussian}(x\mid \mu, \sigma) =
\frac{1}{\sqrt{2\pi \sigma^2}}
\exp{\left(-\frac{1}{2}\frac{(x - \mu)^2}{\sigma^2}\right)}
\end{equation}
This is not the famous Central Limit Theorem, which is related to the
distribution of sums, but can be used to derive it: the distribution of a sum
is a convolution integral, which becomes a product in Fourier-transformed
space, and that product can converge to a Gaussian function by this rule.
Jaynes gives a beautiful alternative proof of the
Central Limit Theorem~\cite{jaynes2003probability}
(Section ``7.16 The central limit theorem'').

@@

% connection to frequentist section
Bayes' theorem is an infamous and necessary result, and it is not seriously
disputed. What has been disputed is whether probability can be applied to
logic, or any application other than the random happenings of events.

### 2022-08-23 main/draft.tex
more probability posteriors etc, ref cartoon

>>>>

infamously arbitrary and subjective. This is reality. They are.
Consider this simple mathematical example with two propositions
$a=\textrm{``}x\textrm{ is even''}$ and
$b=\textrm{``}x\textrm{ equals }2\textrm{''}$.
Clearly $\prob{a}{b} = 1$ since $2$ is indeed even.
But what is $\prob{b}{a}$? We don't know. Perhaps one could argue symmetry
across all even numbers, but there are infinitely many of them so that
probability distribution could not be normalized.

\TODO{\ldots}
